Data Specialization for a Staged Lambda Calculus

### 2015-02-27 paper/implementation.tex
renaming qsStaged

We allow staging annotations around declarations. For example,
\begin{lstlisting}
3`atsigngrnd{`1`
  datatype list = Empty | Cons of int * list
  fun partition (...) = ...
`3`}`1`
fun qsStaged (...) = ...`
\end{lstlisting}
declares the \texttt{list} datatype and \texttt{partition} function at world \bbonep,
by elaborating into:
\begin{lstlisting}
1`val 3`grnd{`Empty3`}` = 3`grnd{`roll (inj ...)3`}`
val 3`grnd{`Cons3`}`  = 3`grnd{`fn (...) => roll (inj ...)3`}`
val 3`grnd{`partition3`}`  = 3`grnd{`fn (...) => ...3`}`
val qsStaged    = fn (...) => ...`
\end{lstlisting}

>>>>

We allow staging annotations around declarations. For example,
\begin{lstlisting}
3`atsigngrnd{`1`
  datatype list = Empty | Cons of int * list
  fun partition (...) = ...
`3`}`1`
fun qss (...) = ...`
\end{lstlisting}
declares the \texttt{list} datatype and \texttt{partition} function at world \bbonep,
by elaborating into:
\begin{lstlisting}
1`val 3`grnd{`Empty3`}` = 3`grnd{`roll (inj ...)3`}`
val 3`grnd{`Cons3`}` = 3`grnd{`fn (...) => roll (inj ...)3`}`
val 3`grnd{`partition3`}` = 3`grnd{`fn (...) => ...3`}`
val qss = fn (...) => ...`
\end{lstlisting}

### 2015-02-27 paper/semantics.tex
overview section is now sec:overview
eliminate inconsistent uses of 'part'ition
renaming qsStaged

\crem{blend in the rest of the text in this section:}
%
\texttt{qsStaged} has a more precise type---it is a first-stage function which
takes a $\curr\mathrm{list}$ (an integer list now) and a $\fut\rmint$ (an
$\rmint$ \emph{in the future}), and returns a $\fut\rmint$.

Indeed, we case on the list as usual, and in the \texttt{Cons} branch,
\texttt{part}ition it. The \texttt{case} expression in \texttt{qSelect} depends
on \texttt{k}, whose type $\fut\rmint$ indicates that it is only available to
second-stage computations. Since \texttt{qsStaged} itself produces a
$\fut\rmint$, the remainder of the function is second-stage code.

The $\next$ wraps a second-stage expression of type $\rmint$ (the \texttt{case}
expression) as a first-stage expression of type $\fut\rmint$ (the result 
of \texttt{qsStaged}). Inside the \texttt{LT} and \texttt{GT} branches, the
$\prev$ unwraps first-stage $\fut\rmint$s (the results of the recursive calls
to \texttt{qsStaged}) as second-stage $\rmint$s. In the \texttt{EQ} branch,
$\pause$ promotes a first-stage $\rmint$ (the head of \texttt{l}) directly to a
second-stage $\rmint$.

@@

Notice that our type system makes it possible to evaluate terms in a staged
fashion, but doing so still requires some machinery, since the stages are
syntactically interleaved within an individual term. For example,
\texttt{qsStaged} contains a first-stage recursive call inside a second-stage
\texttt{case}; thus, an ordinary outermost-first evaluation strategy would not
be properly staged.

@@

A multistage function, such as \texttt{qsStaged} from \ref{sec:examplea}, has
type $A\to \fut (B \to C)$ at \bbonem.%
\footnote{We can rewrite \texttt{qsStaged} in this curried form, or apply a
higher-order currying function.}
To apply this function to arguments $\colmix a A$ and $\coltwo b B$, simply
evaluate the program:
\begin{lstlisting}
2`prev{`1`f a`2`} b`
\end{lstlisting}

We can also directly observe the reuse of first-stage computations across
multiple second-stage computations. To run many order statistics queries
$\texttt{k1},\dots,\texttt{km}$ on the same list, run the program:
\begin{lstlisting}
2`let val qs = prev{`1`qsStaged `3`grnd{`1`[7,4,2,5,9,...,3]`3`}`2`}`
2`in (qs k1,...,qs km)`
\end{lstlisting}
To see that the partial application of \texttt{qsStaged} is only evaluated once,
observe that after the first stage of evaluation, the residual has the form:
\begin{lstlisting}
2`let val yhat = ...`
2`    val qs = yhat`
2`in (qs k1,...,qs km)`
\end{lstlisting}
where $\hat{y}$ is bound to a function containing no first-stage redexes.

>>>>

\crem{blend in the rest of the text in this section:}
%
\texttt{qss} has a more precise type---it is a first-stage function which
takes a $\curr\mathrm{list}$ (an integer list now) and a $\fut\rmint$ (an
$\rmint$ \emph{in the future}), and returns a $\fut\rmint$.

Indeed, we case on the list as usual, and in the \texttt{Cons} branch,
\texttt{partition} it. The \texttt{case} expression in \texttt{qSelect} depends
on \texttt{k}, whose type $\fut\rmint$ indicates that it is only available to
second-stage computations. Since \texttt{qss} itself produces a
$\fut\rmint$, the remainder of the function is second-stage code.

The $\next$ wraps a second-stage expression of type $\rmint$ (the \texttt{case}
expression) as a first-stage expression of type $\fut\rmint$ (the result 
of \texttt{qss}). Inside the \texttt{LT} and \texttt{GT} branches, the
$\prev$ unwraps first-stage $\fut\rmint$s (the results of the recursive calls
to \texttt{qss}) as second-stage $\rmint$s. In the \texttt{EQ} branch,
$\pause$ promotes a first-stage $\rmint$ (the head of \texttt{l}) directly to a
second-stage $\rmint$.

@@

Notice that our type system makes it possible to evaluate terms in a staged
fashion, but doing so still requires some machinery, since the stages are
syntactically interleaved within an individual term. For example,
\texttt{qss} contains a first-stage recursive call inside a second-stage
\texttt{case}; thus, an ordinary outermost-first evaluation strategy would not
be properly staged.

@@

A multistage function, such as \texttt{qss} from \ref{sec:overview}, has
type $A\to \fut (B \to C)$ at \bbonem.%
\footnote{We can rewrite \texttt{qss} in this curried form, or apply a
higher-order currying function.}
To apply this function to arguments $\colmix a A$ and $\coltwo b B$, simply
evaluate the program:
\begin{lstlisting}
2`prev{`1`f a`2`} b`
\end{lstlisting}

We can also directly observe the reuse of first-stage computations across
multiple second-stage computations. To run many order statistics queries
$\texttt{k1},\dots,\texttt{km}$ on the same list, run the program:
\begin{lstlisting}
2`let val qs = prev{`1`qss `3`grnd{`1`[7,4,2,5,9,...,3]`3`}`2`}`
2`in (qs k1,...,qs km)`
\end{lstlisting}
To see that the partial application of \texttt{qss} is only evaluated once,
observe that after the first stage of evaluation, the residual has the form:
\begin{lstlisting}
2`let val yhat = ...`
2`    val qs = yhat`
2`in (qs k1,...,qs km)`
\end{lstlisting}
where $\hat{y}$ is bound to a function containing no first-stage redexes.
